---
title: "Week 6 Exercises"
output: html_notebook
---

```{r setup}
library(tidyverse)
```


# Exercise 1

The data frames df_w and df_f represent repeated measures data from 60 participants. Variables F1-F3 and W1-W3 are "sub-variables" that will be used to make two composite variables F_total and W_total, respectively. 

```{r include=F}
# simulating data
set.seed(1)
df_f <- data.frame(id = rep(1:60, each = 2) %>% as.character,
                 session = rep(2:1, 60),
                 group = rep(1:2, each = 60),
                 F1 = sample.int(4, 120, replace = T) - 1,
                 F2 = sample.int(4, 120, replace = T) - 1,
                 F3 = (sample.int(4, 120, replace = T) - 1)) %>%
  mutate(F3 = F3*(session-1), F2 = F2*(group-1)*(2-session)) %>%
  mutate(group = factor(group))

df_w <- data.frame(id = rep(1:60, each = 2) %>% as.character,
                 session = rep(1:2, 60) %>% factor,
                 group = rep(1:2, each = 60) %>% factor,
                 W1 = sample.int(4, 120, replace = T) - 1,
                 W2 = sample.int(4, 120, replace = T) - 1,
                 W3 = sample.int(4, 120, replace = T) - 1)
```


#### 1.1

Merge the two data frames together.

```{r}

df <- df_f %>%
  mutate(session = factor(session)) %>% # when joining, variable classes must match. therefore session is transformed to factor
  full_join(df_w, by = c("id", "session", "group"))


```

#### 1.2

Using the merged data frame, create the **composite variables F_total and W_total, which are the sums of F1-F3 and W1-W3, respectively** (i.e. their values can range from 0 to 9).

```{r}

df <- df %>% 
  group_by(id, session, group) %>%
  mutate(F_total = sum(c_across(F1:F3)), # or just F1 + F2 + F3
         W_total = sum(c_across(W1:W3))) %>%
  ungroup

df

```



# Exercise 2

#### 2.1

Visualise the distributions of F_total and W_total for the two groups and measurement sessions (for example as boxplots). 

```{r}

df %>%
  gather(var, value, F_total, W_total) %>%
  ggplot(aes(session, value, fill = group)) + 
  geom_boxplot() + 
  facet_wrap(~var) + 
  theme_minimal()

```

#### 2.2

Fit a linear regression model with F_total as the DV, and session and group as predictors.  

```{r}

fit <- lm(F_total ~ session + group, data = df) # a model with main effects
# note that session and group should be factors, not integers!
  
summary(fit)

```

#### 2.3

Look at the means of F_total by group and session. How are they linked to the linear regression model coefficients? 

**Since the model fit has only the main effects for group and session, the coefficients represent the difference in means between groups (2 - 1) and sessions (2 - 1).** 

```{r}
df %>%
  group_by(group) %>%
  summarise(mean_F = mean(F_total)) %>%
  mutate(diff = diff(mean_F)) # diff = coefficient group2

df %>%
  group_by(session) %>%
  summarise(mean_F = mean(F_total)) %>%
  mutate(diff = diff(mean_F)) # diff = coefficient session2

```



# Exercise 3

Visualise the [anscombe dataset](https://rdrr.io/r/datasets/anscombe.html) using ggplot2.

**The key is to get the data into long format, and separate the values (x/y) and sets (1-4) into variables of their own. A few example solutions below.**


```{r}
# method 1
# use pivot_longer and separate set and value simultaneously
anscombe_long <- anscombe %>%
  pivot_longer(everything(), 
               names_to = c(".value", "set"), 
               names_pattern = "(.)(.)") # the names_pattern argument uses regular expressions (inside brackets for each column). The dot . matches any character (except a new line)

```


```{r}
# method 2
# use gather, then separate set and variable, then spread 
anscombe_long <- anscombe %>%
  gather() %>%
  separate(key, into = c("variable", "set"), sep = 1) %>%
  group_by(variable) %>%
  mutate(row = row_number()) %>%
  spread(variable, value) %>%
  select(-row) %>%
  ungroup

```


```{r}
# plot the data
anscombe_long %>%
  ggplot(aes(x, y)) + 
  geom_point() +
  facet_wrap(~set) +
  geom_smooth(method = "lm", se = F) + 
  labs(title = "The anscombe data") +
  theme_bw()

```


