---
title: "Week 1: R Basics"
output: 
  html_notebook:
    toc: yes
---

## Running code, assigning values to objects

You can get output by simply typing into the console, or running (part of) a script file. Output is shown in the console (.R files), or in the editor (.Rmd files, such as this).   
You can run a line of code with the top right button, or Ctrl+Enter (see Tools -> Keyboard Shortcuts Help).  
A chunk in an .Rmd file can be run with the arrow on top of the chunk.  

```{r}
5+5
```


Assign value (10) to object (x) with the arrow.  
You can choose object names quite freely, but it's best to avoid e.g. function names (mean, sd).  
Object names should start with a letter, and can contain letters, numbers, _ and .
See Environment tab in RStudio!  

```{r}
x <- 10

# note: you could just as well use 10 -> x
```

<br/><br/>

Print the object by just typing its name  

```{r}
x

x + 5
```

<br/><br/>

Create a new object based on x  

```{r}
y <- 2*x

```


## Vectors and functions

The basic data type of R is a **vector**, created with *c()*.  
Let's create a vector named "age", with 5 elements:  
```{r}
age <- c(13, 15, 12, 13, 16)

age
```

<br/><br/>

**Functions** in R are of the form *function_name(argument)*  

Some basic functions:  
```{r}
length(age)
mean(age)
```
```{r}
summary(age) 
```

```{r}
str(age) # shows structure of the vector: type, length, first values
```
<br/><br/>

Output from functions can also be saved to objects  

```{r}
mean_age <- mean(age)
```
<br/><br/>

An object can be printed with text using the *paste()* function
```{r}
paste("Mean age:", mean_age)
```
<br/><br/>

Concatenate vectors with *c()*: here, we combine one new element (x) with the age vector
```{r}
new_age <- c(age, x)
length(new_age)
```


#### Accessing the elements of a vector

You can access individual elements of a vector with indices  
```{r}
age[1]
age[3]
age[1:3]
```
<br/><br/>

You can print elements that fulfill a condition.

```{r}
# which ages are above 13?
age > 13
cond <- (age > 13)
age[cond]

```

Above, cond is a **logical vector**, i.e. it consists of TRUE/FALSE values only. These can be used to get subsets of other vectors.  

<br/><br/>

There are also **character (categorical) vectors**.
```{r}
firstname <- c("John", "Mark", "Jane", "Alice", "Sophie")
```

<br/><br/>

Print the structure of the vector
```{r}
str(firstname)

```

## From vectors to data frames

Multiple vectors can be combined into a **data frame**. It will also show up in the Environment tab.  
```{r}
df <- data.frame(age, firstname)

```

```{r}
str(df)
```
<br/><br/>

In Rmd, data frames will be printed neatly as tables, also showing variable types.   
```{r}
df
```
## Data wrangling: base R vs tidyverse
Tidyverse is an alternative library to base R (the traditional style of R code), aiming for more readable and understandable code. Even if base R solutions may sometimes be shorter, they are usually uglier and/or less clear. Below are some examples with both styles. It's good to have an idea of the logic of base R; however, this course will focus on using tidyverse functions for data wrangling.  


#### Base R
Print variable of data frame
```{r}
df$age
```
<br/><br/>

Make new variable (here, centered age values)  
```{r}
df$age_centered <- (df$age - mean(df$age))
```
<br/><br/>

Select variables *age* and *age_centered*
```{r}
df[c(1,3)]
```
<br/><br/>

OR:
```{r}
df[c("age", "age_centered")] 
```
<br/><br/>

Select variables that start with "age":
```{r}
df[grepl("^age", names(df))]  # this uses a thing called regular expressions
```


#### Same in tidyverse
```{r}
library(tidyverse) # load tidyverse
```

<br/><br/>

Make *age_centered* variable with the mutate() function  
```{r}
df <- mutate(df, age_centered = age - mean(age))
```
<br/><br/>

Select variables *age* and *age_centered*  
```{r}
select(df, age, age_centered)
```
<br/><br/>
Select variables that start with "age":
```{r}
select(df, starts_with("age"))
```


##### Pipe operator

Using the **pipe operator %>%** enhances readability of your code. It essentially means that whatever comes *before* the pipe is used as the first argument of the function that comes *after* the pipe.  
For example, variables would be selected like this:  
```{r}
df %>% select(age, age_centered)

```
<br/><br/>

Or an even simpler example:  
```{r}
age %>% mean # instead of mean(age)
```

<br/><br/>

The benefits of the pipe operator become more obvious when we need to carry out many operations in sequence.  The code chunk below:  

1. takes the data frame
2. creates a new variable with *mutate()*  
3. filters rows of the data frame with *filter()*  
4. groups the data frame by gender with *group_by()*  
5. computes mean age by gender with *summarise()*  
(it's common to use line breaks after each pipe)  
```{r}
df %>% 
  mutate(gender = c('Male', 'Male', 'Female', 'Female', 'Female')) %>% 
  filter(age >= 13) %>% # filter ages greater than or equal to 13
  group_by(gender) %>% # group by gender
  summarise(age = mean(age)) # find mean age by gender

```
It's quite easy to understand what is going on by just reading the code (but don't worry if it isn't just yet!), because you can guess what a tidyverse function does based on its name, and the pipe operator helps you to perceive the order of the steps.

### What about figures?

Base R also has its own plotting functions, which are handy for quick and simple visualizations, but the code becomes messy if you want to make your plots look nice (which we will). That is why we won't go deeper into base R plotting during this course, but will instead focus on the **ggplot2** library (part of tidyverse).  

Here's an ugly base R plot that you will learn to make nicer with ggplot2:  
```{r}
plot(df$age, df$age_centered)

```


