---
title: "Week 4 Exercises"
output: html_notebook
---

```{r setup}
library(tidyverse)
```


# Exercise 1

Read the region_scores.csv data (make sure you have the data file in the right folder). 

```{r}
df <- data.table::fread('data/region_scores.csv')

```


Create a figure that shows the distributions (density plots or histograms) of **age** and **score** in separate subplots (facets). What do you need to do first?  

In the figure, set individual x-axis limits for age and score by modifying the `scales` parameter within `facet_wrap()`.

**Hint**: To make things simpler, you can begin by selecting only the variables you need here, i.e. age and score.  

```{r}
df %>%
  select(age, score) %>%
  gather() %>%
  ggplot(aes(value)) + 
  geom_histogram(binwidth = 2) + # use histogram, not geom_bar or geom_col. geom_bar tries to plot counts per distinct value, but we want to plot counts per "bin" (here, 2 units) - this is computed automatically with geom_histogram.
  facet_wrap(~key, scales = "free_x") +
  theme_bw()


# or density:

df %>%
  select(age, score) %>%
  gather() %>%
  ggplot(aes(value)) + 
  geom_density() +
  facet_wrap(~key, scales = "free_x") +
  theme_bw()

```
# Exercise 2

In this exercise, you will use the built-in iris dataset.

```{r}
head(iris)
```


#### 2.1 

Make the data into long format: gather all variables except species into new variables **var** (variable names) and **measure** (numerical values). You should end up with 600 rows and 3 columns (Species, var, and measure). Assign the result into `iris_long`.

```{r}
iris_long <- iris %>%
  gather(var, measure, -Species)
```

#### 2.2

In `iris_long`, separate **var** into two variables: **part** (Sepal/Petal values) and **dim** (Length/Width).  

Then, spread the measurement values to new columns that get their names from **dim**. You must create row numbers by dim group before doing this.  

You should now have 300 rows of variables Species, part, Length and Width (and row numbers). Assign the result into `iris_wide`.

```{r}
iris_wide <- iris_long %>%
  separate(var, into = c('part', 'dim')) %>%
  group_by(dim) %>%
  mutate(row = row_number()) %>%
  ungroup %>%
  spread(dim, measure)
```


#### 2.3

Using `iris_wide`, plot length on the x-axis and width on the y-axis. Colour the points by part.

```{r}
iris_wide %>%
  ggplot(aes(Length, Width, colour = part)) + 
  geom_jitter(alpha = .5) +
  theme_bw()
```