---
title: "Week 3 Exercises"
output: html_notebook
---

```{r setup}
library(data.table)
library(tidyverse)
```


# Exercise 1

#### 1.1

Read the *region_scores.csv* data into R. It should have 240 observations of 6 variables.
```{r}
df <- data.table::fread('data/region_scores.csv')
```

```{r}
summary(df)
```

#### 1.2

Cut the continuous variable **age** into a categorical variable **age_group** that has groups of width 10. Use ggplot2's cutting function.  

Count observations by age group.

```{r}

df <- df %>%
  mutate(age_group = cut_width(age,10)) # use boundary = 0 if you want to start from 0 (in this case, there are no observations for the lowest class, so the first class would be 10-20)

df %>%
  count(age_group)

```


# Exercise 2

The chunk below is supposed to produce a plot but it has some errors.  

The figure should be a scatter plot of **age** on the x-axis and **score** on the y-axis, with points coloured by **education level** (3 levels). It should also have three linear regression lines, one for each of the education levels.  

Fix the code to produce the right figure.  

What happens if you use `geom_jitter()` instead of `geom_point()`?  

**Hint**: Examine the code bit by bit: start by plotting just the scatter plot without `geom_smooth()`, and add the regression lines last.  


```{r}
df %>% 
  ggplot(aes(age, score, colour = factor(education))) + # variables should be inside aes(), and education should be categorical
  geom_jitter() + 
  geom_smooth(method = "lm") +
  labs(colour = "Education", x = "Age", y = "Score") +
  theme_bw()

```

# Exercise 3

#### 3.1

Calculate the mean, standard deviation (sd), and number of observations (n) of **score by education and age group**. Also calculate the standard error of the mean (by using sd and n). Save these into a new data frame (or tibble) named edu_age_stats.  

```{r}
edu_age_stats <- df %>%
  group_by(education, age_group, .drop = FALSE) %>% # .drop = FALSE if you don't want to drop possible missing levels
  summarise(mean_score = mean(score),
            sd_score = sd(score),
            n = n()) %>%
  ungroup() %>%
  mutate(se = sd_score/sqrt(n))  # calculate standard error

edu_age_stats
```

#### 3.2

Using edu_age_stats, plot a bar plot that has **age group** on the x-axis, **mean score** on the y-axis, and **education levels** in subplots (facets).  


Use `geom_errorbar()` to add error bars that represent standard errors of the mean.  


```{r}

edu_age_stats %>%
  ggplot(aes(age_group, mean_score)) + 
  geom_col() +
  geom_errorbar(aes(ymin = mean_score - se, ymax = mean_score + se)) + # error bars centered on the mean
  facet_wrap(~education) +
  theme_bw() +
  labs(x = "Age group", y = "Mean score") +
  scale_x_discrete(labels = c("-25", "26-35", "36-45", "46-55", "56-")) # labels can be changed here, or in the data itself

```

# Exercise 4

#### 4.1

Create a figure that has boxplots of **score** (x-axis) by **region** (y-axis).   

```{r}
df %>%
  ggplot(aes(score, region)) + geom_boxplot() +
  theme_bw() +
  labs(x = "Score", y = "Region")

```
### 4.2

Group the data by region and add mean score by region to a new column **mean_score**. Do this with **`mutate()`** (not `summarise()`). 

Reorder the levels of region based on mean_score.  

**Hint**: Remember to ungroup after creating the mean_score variable.

```{r}
df <- df %>%
  group_by(region) %>%
  mutate(mean_score = mean(score)) %>%
  ungroup %>%
  mutate(region = fct_reorder(region, mean_score))

```


# Exercise 5

Using the data you modified in exercise 4.2, plot **mean scores** (x-axis) by **region** (y-axis) as points. The regions should be ordered by mean score.  

Use `stat_summary()` to add error bars that represent standard errors of the mean.  

**Hint**: Be careful which variable - mean_score or score - you're plotting in each of the geoms.

**The points are calculated using the mean scores, while the error bars are calculated from non-aggregated scores (the summarising is done within stat_summary). Therefore, stat_summary needs to be given different aesthetics.**

```{r}
df %>%
  ggplot(aes(mean_score, region)) + # aesthetics to start with the points
  geom_point() +
  stat_summary(aes(score, region), geom = "errorbar", fun.data = "mean_se", width = .2) + # new aesthetics for error bars
  theme_bw() +
  labs(x = "Mean score", y = "Region")

```


