---
title: "Week 5 Exercises"
output: html_notebook
---

Read the file timeuse_tidy.rds with `readRDS()`. The file contains the dataset that we tidied in the exercise session: records of daily time use from participants over multiple days. Note that since the data has been stored as rds (R-specific format), column types and factor levels are as we left them, and don't need to be re-corrected.

```{r}

```


# Exercise 1

#### 1.1

Create a new variable that contains combined activity classes: "Work or school" (Work, Business, Education), "Free time" (Shopping, Social / Leisure, Home, Vacation), and "Other".  

Calculate the mean time spent on each of the combined activity classes, grouped by **weekday, participant ID, and occ_full_time**.  

```{r}
  
```

#### 1.2 

Visualise the means you calculated.

```{r}

```


# Exercise 2

#### 2.1

What is computed in the code chunk below - what do the numbers tell you?  

Can you think of another way to calculate the same thing?  

```{r}

tidytime %>%
  distinct(indivID, date) %>%
  arrange(date) %>%
  count(date)

```

#### 2.2

Plot the numbers from above (use points, lines, or whatever you think is suitable).  

```{r}

```


# Exercise 3

#### 3.1

Count the **total number of participants** in the data.  

```{r}
  
```

#### 3.2

For each participant, count the **number of separate days** that they recorded their time use on.  

```{r}
  
```


# Exercise 4

Explain step by step what happens in the code chunk below, and what the final figure represents.

```{r fig.height=20, fig.width=10}
tidytime %>%
  group_by(indivID) %>%
  mutate(start_date = min(date)) %>%
  ungroup %>%
  mutate(indivID = factor(indivID),
         indivID = fct_reorder(indivID, start_date) %>% fct_rev()) %>%
  ggplot(aes(date, indivID, colour = month(start_date, label = T))) + 
  geom_line() + 
  geom_point(size=.5, alpha=.1) +
  theme_bw() + 
  scale_y_discrete(breaks = "none") +
  labs(x = "Date", y = "", colour = "Starting month")

```





